\section{RNNs and LSTMs}
\textit{Compare the implementation of Transformers with RNNs/LSTMs in PyTorch.
Provide code snippets for both models and explain why Transformers outperform
RNNs in handling long-range dependencies. Discuss computational trade-offs.}

Recurrent Neural Networks (RNNs) and Long Short-Term Memory
(LSTM) networks were among the earliest neural architectures
designed to handle sequential data.
They process input tokens one step at a time, maintaining
a hidden state that is updated recursively.
In contrast, Transformers rely entirely on attention
mechanisms and process sequences in parallel.
In this section, we compare a Transformer-based sentiment
classifier with an RNN-based classifier implemented in
PyTorch, highlighting architectural differences, performance
characteristics, and computational trade-offs.

The RNN-based sentiment classifier processes a sequence
incrementally, updating its hidden state at each time step.
Given an input sequence $(x_1, \dots, x_T)$, the hidden state
$h_t$ is computed as
\begin{equation}
h_t = f(h_{t-1}, x_t),
\end{equation}
where $f$ denotes the recurrent update function.
In the case of LSTMs, this update includes gating mechanisms
that regulate information flow through input, forget, and
output gates, allowing the model to mitigate the vanishing
gradient problem.
After processing the full sequence, the final hidden state
(or a pooled representation of all hidden states) is passed
to a linear classifier to predict the sentiment label.
The RNN classifier implementation is shown in
\Cref{lst:rnn_classifier}.
\input{listings/rnn.tex}

While LSTMs improve upon vanilla RNNs, they still suffer
from inherent limitations when modeling long-range
dependencies.
Information must propagate sequentially through many time
steps, making it difficult to preserve context over long
sequences.
This sequential dependency also prevents parallelization
across time steps, resulting in slower training and inference.

The Transformer-based classifier, shown previously in
\Cref{lst:transformer}, addresses these limitations by using
self-attention instead of recurrence.
Rather than relying on a single hidden state that is updated
over time, self-attention allows each token to directly
attend to all other tokens in the sequence.
As a result, long-range dependencies can be modeled more
effectively, regardless of the distance between tokens.
Additionally, the absence of recurrence enables full
parallelization over sequence positions, significantly
reducing training time on modern hardware.

From a computational perspective, the two architectures
exhibit different trade-offs.
RNNs and LSTMs have linear complexity with respect to
sequence length, making them memory-efficient for long
sequences.
However, their lack of parallelism limits throughput and
scalability.
Transformers, on the other hand, incur a quadratic
computational cost due to the self-attention mechanism,
as every token attends to every other token.
This increases memory usage and can become a bottleneck
for very long sequences.

Despite this higher theoretical complexity, Transformers
generally outperform RNNs and LSTMs in practice for sentiment
classification and other NLP tasks.
Their ability to model global context, combined with efficient
parallel computation on GPUs, leads to faster convergence and
better empirical performance.
As a result, Transformer-based models have largely replaced
recurrent architectures as the dominant approach for handling
long-range dependencies in modern sequence modeling tasks.
