\section[Sequence-to-Sequence Model]{Sequence-to-Sequence Machine Translation Model}\label{sec:Sequence-to-Sequence Model} % (fold)
\textit{Implement a Transformer-based sequence-to-sequence model in PyTorch for 
machine translation (English → French). Explain the encoder-decoder structure,
attention mechanism, and training process. Provide code snippets and discuss 
evaluation metrics such as BLEU score}

The full working implementation is available in \mintinline{text}{code/seq2seq.py}. 
Only important parts of the code are discussed below to
provide an illustration of the working.
The code largely follows the flow outlined in the PyTorch 
Documentation~\cite{robertsonNLPScratchTranslation}.
The dataset was downloaded from Tatoeba~\cite{TatoebaCollectionSentencesa},
and preprocessed before usage. 
Preprocessing included extracting pairs of English sentences with French
translations, and converting Unicode strings into ASCII strings. 

First, we define a class to help build a vocabulary and 
get one-hot encoded vectors.
The \mintinline{py}{class Lang} is responsible for
the ingestion of sentences and the storage of dictionaries
holding the indices of words present in the language.
A simple tokenizer is used, and tokens are split by whitespaces.

The encoder-decoder architecture is one of three possible
transformer architectures~\cite{aitkenUnderstandingHowEncoderDecoder2021}. 
It is especially suited for summarization and machine translation
as there is no restriction on both the input and output length.
It uses the encoder part to calculate a dense context vector that
effectively captures all the data from the input vector. 
The encoder we have used is an RNN with a Gated Recurrent Unit
that allows it to have better context and memory.
The encoder code is given in \Cref{lst:seq2seq_encoder}
\input{listings/seq2seq_encoder.tex}


The decoder part of the transformer consists of another RNN-GRU,
and uses attention along with already generated tokens to
get a probability distribution of possible continuations.
The decoder can work in two ways, by using previous tokens as mentioned earlier,
or by ``teacher-forcing'' where we can force the next step of attention to use
tokens provided by a ground truth. 
Teacher-forcing allows a model to train faster and make less mistakes by not
allowing mistakes to propagate far into the model.
Code for the decoder can be found in \Cref{lst:seq2seq_decoder}
\input{listings/seq2seq_decoder.tex}

Attention is a mechanism to control the tokens 
that a model considers relevant and important. 
Attention allows a decoder access to all the states
of the encoder, and selectively picks the most important ones.

For every generated token, attention calculates an alignment
score, called the \emph{attention weight}.
Instead of compressing all information from the encoder into
a single fixed-length vector, attention computes a weighted
combination of all encoder hidden states.
This allows the decoder to dynamically focus on different
parts of the input sequence at each decoding step.

In this work, we specifically use \emph{Bahdanau attention}~\cite{bahdanauEndtoendAttentionbasedLarge2016},
which is an additive attention mechanism.
Given the encoder hidden states $h_i$ and the current decoder
hidden state $s_t$, an alignment score is computed as
\begin{equation}
e_{t,i} = v_a^\top \tanh(W_a s_t + U_a h_i),
\end{equation}
where $W_a$, $U_a$, and $v_a$ are learnable parameters.
These scores are normalized using a softmax function to obtain
attention weights
\begin{equation}
\alpha_{t,i} = \frac{\exp(e_{t,i})}{\sum_j \exp(e_{t,j})}.
\end{equation}
The context vector $c_t$ is then computed as a weighted sum
of encoder hidden states
\begin{equation}
c_t = \sum_i \alpha_{t,i} h_i.
\end{equation}
This context vector is concatenated with the decoder hidden
state and passed to the output layer to generate the next token.
The attention implementation used in this project is shown in
\Cref{lst:seq2seq_attention}.
\input{listings/seq2seq_attention.tex}

The training process consists of minimizing the negative
log-likelihood loss between the predicted output sequence
and the ground truth translation.
Given a target sentence $y = (y_1, \dots, y_T)$, the loss is defined as
\begin{equation}
\mathcal{L} = -\sum_{t=1}^{T} \log p(y_t \mid y_{<t}, x),
\end{equation}
where $x$ is the input English sentence.
The model is trained using the Adam optimizer, and gradients
are computed using backpropagation through time.
Teacher-forcing is applied with a fixed probability, allowing
the decoder to alternate between using its own predictions
and ground-truth tokens during training.
This improves convergence speed and stability.

During evaluation, the model is switched to inference mode
and generates translations autoregressively without
teacher-forcing.
To quantitatively assess translation quality, we use the
BLEU (Bilingual Evaluation Understudy) score, which measures
the overlap of $n$-grams between the generated translation
and a reference translation.
BLEU is defined as
\begin{equation}
\text{BLEU} = \text{BP} \cdot \exp\left( \sum_{n=1}^{N} w_n \log p_n \right),
\end{equation}
where $p_n$ denotes the precision of $n$-grams, $w_n$ are
uniform weights, and BP is a brevity penalty that discourages
overly short translations.
BLEU provides a standardized and widely-used metric for
comparing machine translation systems, and is well-suited
for evaluating the English-to-French task considered here.
