\section{Sinusoidal Positional Encoding}\label{sec:sinusoidal_positional_encoding} 
\textit{Discuss the importance of positional encoding in Transformers. Implement
sinusoidal positional encoding in PyTorch and demonstrate how it integrates with
input embeddings. Explain why positional information is critical for sequence
modeling.}

Transformers process input sequences in parallel and do not
contain any inherent notion of token order.
As a result, positional information must be explicitly
provided to the model in order to preserve the structure
of sequential data.
Without positional encoding, a Transformer would treat
the input as a bag of tokens, making it incapable of
distinguishing between different permutations of the same
sequence.
Positional encoding therefore plays a critical role in
sequence modeling tasks such as machine translation,
language modeling, and text summarization.

The original Transformer architecture introduces
\emph{sinusoidal positional encodings}~\cite{vaswaniAttentionAllYou2017},
which inject position-dependent signals into the input
embeddings using deterministic functions.
These encodings are added directly to the word embeddings,
allowing the model to leverage both semantic and positional
information simultaneously.
Since the positional encodings have the same dimensionality
as the embeddings, this integration does not increase
the model complexity.

For a given position $pos$ and embedding dimension $i$,
the sinusoidal positional encoding is defined as
\begin{equation}
\text{PE}(pos, 2i) = \sin\left(\frac{pos}{10000^{2i / d_{\text{model}}}}\right),
\end{equation}
\begin{equation}
\text{PE}(pos, 2i + 1) = \cos\left(\frac{pos}{10000^{2i / d_{\text{model}}}}\right),
\end{equation}
where $d_{\text{model}}$ denotes the embedding dimension.
Sine functions are used for even indices and cosine
functions for odd indices, resulting in a unique positional
encoding for each position.
This formulation ensures that positions that are close in
the sequence have similar encodings, while still allowing
the model to distinguish distant positions.

An important property of sinusoidal positional encodings
is that they allow the model to generalize to sequence
lengths longer than those seen during training.
Since the encodings are generated using fixed mathematical
functions rather than learned parameters, they can be
computed for arbitrary sequence lengths without retraining.
Additionally, relative positions can be expressed as linear
functions of these encodings, which is beneficial for
attention-based models.

In practice, positional encodings are added element-wise
to the input embeddings before being passed to the
Transformer encoder.
If $E \in \mathbb{R}^{T \times d_{\text{model}}}$ denotes
the token embeddings and $P \in \mathbb{R}^{T \times d_{\text{model}}}$
the positional encodings, the combined input is given by
\begin{equation}
X = E + P.
\end{equation}
This simple operation allows every attention layer in
the network to access positional information implicitly.

A PyTorch implementation of sinusoidal positional encoding
is shown in \Cref{lst:positional_encoding}.
The code constructs a positional encoding matrix once and
registers it as a buffer, ensuring that it is not treated
as a learnable parameter.
This positional encoding is then added to the input
embeddings at each forward pass, providing
positional information into the Transformer pipeline.
\input{listings/sinpoe}
